# Functions and Data Types

```{r 02-setup, echo = FALSE}
knitr::opts_chunk$set(fig.align = "center", warning = FALSE, message = FALSE, fig.height = 7, fig.width = 12, cache = TRUE)
```

<a name = 'functions'></a>

## Functions

A <span class = 'vocab'>**function**</span> in `R` is a set of steps, operations, and procedures that are done to data in a specific order. `R` has some functions that are built into the language (many of which we'll go through in this book), but you are also able to write your own functions as well.

Functions take <span class = 'vocab'>**arguments**</span>, which are data points and other items that the function needs to do its job. Think of them like a variable, where you can change the value that they take each time you need to run the function. These can be anything from data points themselves, to colors and sizes for plots, to even other functions if necessary. After performing the steps and calculations that they're supposed to do, they <span class = 'vocab'>**return**</span>, or give out, the information.

<a name = 'default-argument-note'></a>

<span class = 'note'>*Note: in *</span><span style = 'color: #a21bd7'>`R`</span><span class = 'note'>*, arguments of functions may have pre-defined values. In this case, unless you specify differently when you*</span> <span class = 'vocab'>call</span> <span class = 'note'>*(use) the function, this pre-defined, or*</span> <span class = 'vocab'>default</span><span class = 'note'>*, value will be used instead.*</span>

To write your own functions, you need to make use of the `function()` function. You give your function a name, then specify its arguments as the arguments in `function()`, and include your steps inside of a set of curly braces (`{}`). To tell `R` what to return as the output of your new function, you have two options: you can either just leave it as the last line inside of the curly braces, or you can explicitly state it inside the parenthesis of `return()`.

We'll put a simple example of a <span class = 'vocab'>user-defined function</span> here to illustrate how simple and useful writing a function can be, although you may not completely understand what's going on right now. And **that's okay, and honestly that's *expected* at this point.** We haven't covered what's going on here (it's only Chapter 2!), but if we don't introduce functions conceptually, we can't refer to and/or write and teach them them as we go through the book. They're helpful tools that can save you a *ton* of time as you get better in `R`.

The function we're going to build is called `doubler`. It will take one argument, `x`, and return whatever the double of `x` is. See if you can match the parts in this function to the process we just outlined and with the information about `doubler()` that we just gave you!

```{r 02-doubler-function}
doubler = function(x){
  2 * x
}
```

Then, to call the function, you simply put the name of the function, then without a space, put a set of parenthesis. Inside of these parenthesis, specify the arguments required to make the function run properly. Here's an example of how to call the `doubler()` function we just wrote:

```{r 02-doubler-examples}
doubler(x = 2)
doubler(4)
doubler(100)
```

As you can see, `x` can be any number, and `doubler()` just takes the number (`x`) and doubles it.

<span class = 'tip'>*Pro tip: you don't always need to specify the name of an argument. In the second example of using*</span> <span style = 'color: #00ce45'>`doubler()`</span><span class = 'tip'>,</span> <span style = 'color: #00ce45'>`R`</span> <span class = 'tip'>*interprets*</span> <span style = 'color: #00ce45'>`4`</span> <span class = 'tip'>*to be what*</span> <span style = 'color: #00ce45'>`x`</span> <span class = 'tip'>*is supposed to be. When there's more than one argument needed for a function, you can either give them in the same order the function looks for them (which you'll learn about [here](#helpfiles)), or you can specify them by name.*</span>

## The Types {#types}

Not all data is of the same <span class = 'vocab'>type</span>, or usage format. What this means is that different kinds of information from a dataset get evaluated differently in `R`. To check what type a piece of data is, you can use the `class()` function. Let's go through a few of the most common types of data:

- `numeric`: Numeric data is data that is *only* numbers. These can be positive, negative, 0, decimals, or even infinity ($\infty$)

<a name = '02-character'></a>

- `character`: Character data is anything that involves a letter or special character. These will be denoted by `''` (single quotation marks) or `""` (double quotation marks). Characters are also called <span class = 'vocab'>strings</span>. It's important to note that `2` is of type `numeric`, while `'2'` is of type `character`. A quick check using the `class()` function:

```{r 02-type-check}
class(2)
class('2')
```

<a name = 'logical'></a>

- `logical`: Logical data, also known as <span class = 'vocab'>boolean data</span>, is just a series of `TRUE` or `FALSE` values. While this may not necessarily seem like the most useful form of data right now, it's important to know that this type of data exists. `R` evaluates `T` to be `TRUE` and `F` to be `FALSE`, so it's equally valid to use `T` and `F` in place of `TRUE` and `FALSE`, but it's better practice to use `TRUE` and `FALSE` since we may want to use `T` and `F` as variables. More on this in a little

- `factor`: Factor data is simply categories. This type of data is really useful for later when we want to split the information on variables such as gender, location, or a variety of other categorical features in the data

<a name = 'na'></a>

- `NA`: This isn't actually its own type of data, but it represents a missing value. These can become pesky, but there are ways to work around them. We can choose to replace them with `0` or any other value we want, we can ignore them in our computations, or we can do something completely different with them altogether. The important thing to remember about `NA` values is that they exist and should be acknowledged.

## Vectors, Lists, and Data Frames

Each of the data types listed [above](#types) describes a single point of data, called a <span class = 'vocab'>scalar</span>. However, we usually we don't have data given to us as one-by-one pieces of information. We're normally given whole datasets at a time, or at least groups of related data, and they're much easier to work with.

### Vectors {-}

<a name = 'vectors'></a>

A <span class = 'vocab'>vector**</span> is a grouped set of data. Think about it as if it were the answer to a *single* question from a survey from all students in the class, or the heights of all basketball players in the NBA. We've actually been working with vectors all along! We'll discuss it more in [chapter 4](#Vectorization).

`R` actually treats every value as a vector. That's why, as you may have noticed, lines of output begin with `[1]`. This indicates the <span class = 'vocab'>index**</span> (position) in the vector that is at the start of the line. Any time we've had any type of data, `R` has just treated it as a vector of length 1.

One important thing to note about vectors is that all members of the vector must be of the same type. If they aren't, they will be <span class = 'vocab'>coerced**</span> (changed) to be of the same type. To create a vector, we can use the `c()` function. This function combines the <span class = 'vocab'>elements**</span> (individual data points) and turns them into a vector. Separate the parts with a comma (`,`). Here's a few examples:

```{r 02-vector-examples}
c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

c(TRUE, FALSE, T, F, T, FALSE)

c('R', 'is', 'fun')

c(1, 'apple', 2, 'banana')
```

Note how in the last example, everything appears inside of double quotation marks. This is an example of coersion in action. `1` and `2` are recognized as type `integer`, and `'apple'` and `'banana'` are recognized as type `character`. Since it's easier to change a number to a character than it is to go the other way, `1` and `2` become characters.

It may seem cumbersome, time-consuming, and tedious to type out numbers in order as we did in the first vector. `:` to the rescue! Another way that we can create that vector is by putting the first number we'd like in our vector on the left side of the `:`, and the last number on the right.

```{r 02-colon}
c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

1:10
```

As you can see, they both produce the same output. Let's say we wanted to only include even numbers. Luckily, there's a function that allows us to do that as well, and that's the `seq()` function. To use it, we start by putting the first number in our vector, then the last number we want in the vector, and finally the amount we want to increment by in each place. To get the even numbers between 1 and 10, we want the following sequence:

```{r 02-seq}
seq(2, 10, 2)
```

Now, let's say that we want to actually get at the contents of a part of a vector. We can accesss it by using its index. Unlike some other languages (Python, for example), `R` starts its indexing at 1, not 0. If we want to access the 3$^{\text{rd}}$ element of the vector we just created, use the index of the part we're interested in (3) and put it inside of a single set of square brackets (`[]`) next to the vector. This will return the element in that location.

```{r 02-index}
seq(2, 10, 2)[3]
```

To figure out how many elements are contained in a vector, we can use the `length()` function. This information is also displayed in the `Environment` tab we configured in [chapter 1](#env), assuming that the vector is stored as something in the global environment.

The last function we should mention here is the `rep()` function. Similar to `seq()`, this function allows you to create a vector of the same number, repeated any number of times. Want the number 10 to be repeated 30 times? `rep()` makes this easy, as you can write `rep(10, 30)`. The first argument is the number or vector you'd like to repeat, and the second is the number of times you'd like to repeat it. If the first argument is a vector, and you'd like to repeat each element a certain number of times, include the `each = ` argument, with the number of times that you'd like each element to repeat. Let's see `rep()` in action:

```{r 02-rep-examples}
rep(2, 10)
rep(c(1, 2, 3), 3)
rep(c(1, 2, 3), 3, each = 2)
```

### Lists {-}

<a name = 'lists'></a>

On the surface, there's not much difference between a list and a vector. The biggest difference is that a list can contain different types of data, whereas a vector cannot. To create a list, we can simply use the `list()` function, again putting all the different parts we want included inside the parenthesis, separated by a comma.

```{r 02-list-ex}
list(1:10)

list(1, 'apple', 2, 'banana')
```

Lists can be made up of vectors as well. That is, each element of a list is able to be a vector, since a list doesn't care what type of data each of its elements is. To access a list's elements, we want to use a double set of square brackets (`[[]]`) with the index we'd like to access.

```{r 02-list-ex-2}
list(c('apple', 'banana'), c(1.25, 2.50), 3)[[2]]
```

The above example also illustrates that lists don't need all list elements to be of the same length. Note that the third element of the list is only of length 1 (it's just the number 3), but the other two elements are of length 2.

Lastly, lists are able to have named elements. To name an element, all you have to do is type `NAME OF ELEMENT = ` before each element, where `NAME OF ELEMENT` is whatever name you'd like to assign it. In the above example, let's say we wanted to call the first element `fruits`, the second element `prices`, and the third element `aisle`. Then, our list would look like this:

```{r 02-list-named}
list(fruits = c('apple', 'banana'), prices = c(1.25, 2.50), aisle = 3)
```

We can then use a `$` to go into the list and "pull out" that element (the vector with the corresponding name). We can then use vector indexing rules to get a particular element from the vector. If we wanted to get `'banana'` from our list above, we have two options.

```{r 02-get-banana}
list(fruits = c('apple', 'banana'), prices = c(1.25, 2.50), aisle = 3)$fruits[2]
list(fruits = c('apple', 'banana'), prices = c(1.25, 2.50), aisle = 3)[[1]][2]
```

As you can see, both options return `'banana'`, so these options are equivalent.

### Data Frames {-}

The last major type of combined data storing we need to talk about is a <span class = 'vocab'>data frame</span>. You can think of a data frame as a big table with the data you'd like. Each row of data is called an <span class = 'vocab'>observation</span>, and each column represents a <span class = 'vocab'>feature</span> or a variable. We'll use a few of our own data frames throughout the semester, but it's good to know that `R` comes with some of its own data frames already. This data comes from the `crabs` data frame in the `MASS` package (see <a href = 'packages-and-getting-help.html#Packages'>chapter 3</a> for more information on packages). The first 6 rows are shown below.

```{r 02-data-frame-example, echo = FALSE}
library(MASS)
knitr::kable(head(crabs), format = 'html', align = 'c')
```

To see a data frame, you'll want to use the `View()` command. This will open the data frame in the [`Source`](#src) pane.

A few things about data frames:

- They're really just an easy-to-see list. You can access any column (feature) by using the `$` operator. The <span class = 'vocab'>syntax</span> (way to write the code) is: `df_name$column_name`, where `df_name` and `column_name` are the data frame name and column name respectively

- All columns (or list elements) must be of the same length. They may contain `NA` values, but their lengths *must* be the same

- To get the number of rows of a data frame, use the `nrow()` function. To get the number of columns, you can either use `length()` (since, as stated before, it's just a list of vectors), or `ncol()`. This information is also in the `Environment` tab.

- You can make your own with the `data.frame()` function. Just put the vectors you'd like to include, separated by commas, inside of the parenthesis. Just like with lists, you can name the columns of a data frame as you create it. As long as the vectors are of the same length, you'll be making data frames in no time!

<a name = 'coercion'></a>

## Coercing To Other Types

The last point we'll make about different kinds of data is that you can coerce it yourself to be of another type. There are a lot of functions, the `as._()` functions, that are helpful here. Have a character string that's just a number? No problem! we saw before that `class('2')` was of type `character`. What about if we wanted it to be of type `numeric`?

```{r 02-coersion-ex}
class(as.numeric('2'))
```

Awesome.

## `identical()`

This is as good a time as any to introduce the `identical()` function. What this does is checks if the things supplied to it are the identical. It returns `TRUE` if the arguments are identical, and `FALSE` if they're different. Examples, with some of the syntax described above, are as follows:

```{r 02-identical-examples}
identical(1, 1)
x = 1:10 # This has length 10
y = 2:10 # This has length 9
identical(x, y)
identical(x[10], y[9])
```

This function is very helpful when you want to check if two vectors or lists have the same information. It's also particularly useful when you want to check if the outputs or results of different functions are the same if you're reorganizing/rewriting code.
