# Loading Data

Like we learned <a href = 'downloading-r-and-rstudio.html#env'>before</a>, you'll be able to see what data, variables, and other things you've loaded and stored in `R` by checking your `Environment` tab. Right now, yours is most likely empty, and that's okay! We've learned a lot, but haven't started doing a ton yet. We will soon, but first we need to learn how to get data into `R` so that we can start exploring it.

## Data Files

Most of the time, you'll want to work with something called a <span class = 'vocab'>CSV</span> file. CSV stands for **C**omma **S**eparated **V**alue, which means that all of the data within an observation is separated by a comma [Here's an example CSV file](data/Combined Fall 2017 Survey 1.csv) that you can download. To see the raw file, you can open it in a program like TextEdit on a Mac or Notepad on PC. Usually, a CSV file will default to open in a spreadsheet program like Excel since it's easy to <span class = 'vocab'>parse</span> (read in and break up) into columns -- this is because of the commas!

## Reading in CSV Files

To read a CSV into `R`, there's a great function called `read.csv()`. It will read in your file as a data frame. To read in the file, you just need to put the name of the file as a charachter string (remember your <a href = 'functions-and-data-types.html#02-character'>data types</a>?) in between the parenthesis. This is the only required <span class = 'vocab'>argument</span>, or input to a function, that you need to supply, however there are a few others to note:

- The `stringsAsFactors` argument can be either a `TRUE` or `FALSE` value. When `TRUE` (this argument's default value), it takes any character strings in the data and coerces them to be factors. Sometimes this may be okay, sometimes not

- The `header` argument. This indicates whether or not a header row is present in the CSV file, which would contain names for all of the columns. It defaults to `TRUE`, but it's a good idea to double check your data and make sure that one's present. You can check the names and rename if you wish by using the `names()` function discussed in <a href = 'variables.html#naming-lists-and-data-frames'>chapter 3</a>

### Directories {-}

One common issue is the location on your computer where the CSV is located. If you just type the name of the file as the argument for `read.csv()`, `R` will look for it in your <span class = 'vocab'>working directory</span>, or default file lookup location. However, if your file isn't present in the working directory, you'll likely get an error message. If this is the case, you have a few options:

1) You can supply the full filepath to the data as the string. Watch [this video](https://www.youtube.com/watch?v=kIhGavBqXYc) to find how to find the full filepath for Mac or [this one](https://www.youtube.com/watch?v=th23yuZ9fOw) for PC. While it may be more work to find, it'll guarantee that you import the right file.

2) You can change your working directory with the `setwd()` function, supplying the path to the directory as a `character` string argument. This still requires knowing where your file is located, but if you plan to work with multiple files, this isn't a bad option. You can check your current working directory with the `getwd()` function without any arguments.

<a name = 'loading-example'></a>

### Loading Example {-}

Now let's import that example CSV file. It's the combined results from Survey 1 of STAT 100 and STAT 200. Remember, we should name the new data frame so that we can look at it and refer back to it. We'll use the CSV file `'Combined Fall 2017 Survey 1.csv'`, which does have a header row, and is located in a subdirectory of our working directory called `data` (this is the directory that holds all of our data files for the book). It's in a directory called `data`, has a header row, and we'll keep `character` variables as `characters`. When we import it, we'll call it `survey1`:

```{r 06-import-ex}
survey1 = read.csv('data/Combined Fall 2017 Survey 1.csv', header = TRUE, stringsAsFactors = FALSE)
```

To see the first few observations, you can use the `head()` function, passing the data frame's name (in our case, `survey1`). This will display the observations in your `Console`, and will look like this:

```{r 06-head-loaded-data}
head(survey1)
```

Use the `View()` function, again passing the name of the data frame as the argument. It'll display much cleaner and clearer in your `Source` pane, looking more like this:

```{r 06-preview-loaded-data, echo = FALSE}
knitr::kable(head(survey1), align = 'c', format = 'html', booktabs = TRUE)
```

Much better.

## Writing Files

After you finish with your analysis, you may wish to save the data frame(s) that you've created. Similar to the `read.csv()` function that allows you to import a CSV, the `write.csv()` function will allow you to write your own CSV files to your computer to save and send as needed.
