---
title: "STAT 100 Honors Project"
author: "[Ross Drucker](mailto:ross.a.drucker@gmail.com)"
site: bookdown::bookdown_site
output: bookdown::gitbook
knit: bookdown::render_book
documentclass: book
favicon: img/favicon.png
#bibliography: [book.bib, packages.bib]
#biblio-style: apalike
link-citations: yes
highlight: haddock
github-repo: rstudio/bookdown-demo
---
# Preface {-}

```{r index-setup, echo = FALSE}
knitr::opts_chunk$set(fig.align = "center", comment = '', warning = FALSE, message = FALSE, fig.height = 7, fig.width = 12, cache = TRUE)
library(magrittr)
```

Welcome to the STAT 100 Honors Project! Make sure that you've gotten your HCLA approved and submitted to your college.

## Overview {-}

This book is meant to run parallel to the [STAT 100](http://courses.las.illinois.edu/stat/stat100/) course notes. The hope is that STAT 100 students learn how the concepts they learned are put to use in modern statistical settings. Throughout, students will not only gain familiarity with the concepts in the notebook, but will also gain an understanding on how to use the `R` programming language to solve every-day problems in the real world.

## Conventions {-}

A few things to note throughout this book:

- Code that is written in `monospace` font refers to either RStudio keyboard shortcuts or R code, and it will be syntax-highlighted when appropriate

```{r index-test, eval = FALSE}
1 + 1
a = 1
b = 2
a + b
sqrt(4)
```

- Output will begin with `##` on the side

- Vocab words will appear like <span class = 'vocab'>this</span> throughout the text

- <span class = 'tip'>*Helpful hints ("Pro tips") will appear in this color and be in italics. They are usually shortcuts to use in*</span> <span style = 'color: #00ce45'>`R`</span> <span class = 'tip'>*to help make code writing and editing easier to do*</span>

- <span class = 'note'>*Things to keep in mind or pay attention to will look like this. They'll usually begin "Note:"*</span>