---
title: "STAT 100 Honors Project"
author: "[Ross Drucker](mailto:ross.a.drucker@gmail.com)"
site: bookdown::bookdown_site
output: bookdown::gitbook
knit: bookdown::render_book
documentclass: book
favicon: img/favicon.png
#bibliography: [book.bib, packages.bib]
#biblio-style: apalike
link-citations: yes
highlight: haddock
github-repo: rossdrucker/STAT-100-Honors-Project
---
# Intro {-}

```{r index-setup, echo = FALSE}
knitr::opts_chunk$set(fig.align = "center", comment = '', warning = FALSE, message = FALSE, fig.height = 7, fig.width = 12, cache = TRUE)
library(magrittr)
```


**The who**: This project is available to James Scholars interested in learning the statistical computing language `R`. Please sign up ASAP because this is a pilot project and enrollment is limited to 50 students.

**The what**: Our James Scholar Honors Project teaches you to use the statistical software, `R`. Be prepared to spend approximately 2 hours per week on this project.

**The when**: Sign up ASAP by getting a HCLA form from your dean/advisor. As soon as the paperwork goes through, you're ready to go! The deadline to sign up is **Monday, January 29^th^ at 11:59pm**. There will be weekly assignments for the remainder of the semester.

**The how**: The project consists of weekly notes to read ($\approx$1 hour) and weekly assignments to do on Lon-Capa ($\approx$1 hour). You need to complete all the assignments and turn them in before the due date each week. Late assignments will not be accepted on Lon Capa, so make sure to keep a close eye on the calendar. All assignments are due on Sundays at 11:59pm. The first one is due **Sunday February 4^th^, 2018**. It will be posted on Friday January 19th.

**The where**: No need to leave home and get out of your PJs for this one. All notes and assignments can be found online!

**The why**: Why not!? It's a great way to get your James Scholar credit, as well as to learn a useful program in the process. Knowledge of R is a very marketable skill. Also, the project is in sync with class. You'll be able to see how what you learn in class can be done in `R`. This will help reinforce the concepts that you need to know to succeed in Stat 100 and give you an introduction to programming

## About This Book {-}

This book is meant to run parallel to the [STAT 100](http://courses.las.illinois.edu/stat/stat100/) course notes. The hope is that STAT 100 students learn how the concepts they learned are put to use in modern statistical settings. Throughout, students will not only gain familiarity with the concepts in the notebook, but will also gain an understanding on how to use the `R` programming language to solve every-day problems in the real world.

## Conventions For This Book {-}

A few things to note throughout this book:

- Code that is written in `monospace` font refers to either RStudio keyboard shortcuts or R code, and it will be syntax-highlighted when appropriate

```{r index-test, eval = FALSE}
1 + 1
a = 1
b = 2
a + b
sqrt(4)
```

- Output will begin with `##` on the side

- Vocab words will appear like <span class = 'vocab'>this</span> throughout the text

- <span class = 'tip'>Helpful hints ("Pro tips") will appear in this color and be in italics. They are usually shortcuts to use in</span> <span style = 'color: #00ce45'>`R`</span> <span class = 'tip'>to help make code writing and editing easier to do</span>

- <span class = 'note'>Things to keep in mind or pay attention to will look like this. They'll usually begin "Note:"</span>
